### 2024-01-26 misc/expo/additive.tex
Hazem created

\edef\lempetridis{\the\thmcount}
\parenproclaim Lemma {\advthm} (Petridis, {\rm 2012}). Let $A$ and $B$ be subsets of an abelian
group and let $K$ be such that $|A_0+B| = K_0|A_0|$. Then there exist $A\subseteq A_0$ and
$K\le K_0$ such that $|A+B+C| = K|A+C|$ for every finite subset $C$ of the group.

>>>>

\edef\lempetridis{\the\thmcount}
\parenproclaim Lemma {\advthm} (Petridis, {\rm 2012}). Let $A$ and $B$ be subsets of an abelian
group and let $K_0$ be such that $|A_0+B| = K_0|A_0|$. Then there exist $A\subseteq A_0$ and
$K\le K_0$ such that $|A+B+C| \le K|A+C|$ for every finite subset $C$ of the group.

### 2024-01-26 misc/expo/entropy.tex
Hazem created

\def\eps{\epsilon}
\def\FF{{\bf F}}
\def\bar#1{\overline{#1}}
\def\hat#1{\widehat{#1}}
\def\norm#1{|\!|#1|\!|}
\def\bignorm#1{\big|\!\big|#1\big|\!\big|}
\def\Norm#1{\Big|\!\Big|#1\Big|\!\Big|}
\def\normm#1{\bigg|\!\bigg|#1\bigg|\!\bigg|}
\def\Bohr{\op{\rm Bohr}}
\def\Eta{\op{\bf H}}
\def\II{\op{\bf I}}
\def\given{\mathbin{|}}

\widemargins
\bookheader{ENTROPY AND THE FREIMAN--RUZSA THEOREM}{MARCEL K. GOH}

\maketitle{Entropy and additive combinatorics}{by}{Marcel K. Goh}{23 November 2023}

@@

\proclaim Let $X$ and $Y$ be independent random variables. Then
$\Eta\{Y\given X\} = \Eta\{Y\}$ and consequently
$\Eta\{X,Y\} = \Eta\{X\} + \Eta\{Y\}$.

\proof Suppose $X$ takes values in a finite set $A$. Then for all $x\in A$, the distribution of
$Y$ and $Y$ given that $X = x$ is the same, so
$$\Eta\{Y \given X\} = \sum_{x\in A} \pr\{X = x\} \Eta\{Y\given X = x\} = \sum_{x\in A} \pr\{X = x\} \Eta\{Y\}
=\Eta\{Y\}.$$
The second version of the statement is a consequence of the additivity axiom.\slug

@@

\proclaim Corollary \advthm. For a discrete random variable $X$, we have $\Eta\{X^n\} = n\Eta\{X\}$.

@@

In the case where $X$ is the uniform distribution on $\{0,1\}$, and $X_1,\ldots,X_n$ are independent
and distributed as $X$,
we can additionally use the normalisation axiom to conclude that
$$\Eta\{X^n\} = \Eta\{X_1,\ldots,X_n\} = \sum_{i=1}^n \Eta\{X_i\} = \log_2(e)\cdot n,$$
by induction. For general random variables $X_1,\ldots,X_n$, induction also yields the {\it chain rule}
$$\Eta\{X_1,\ldots,X_n\} = \Eta\{X_1\} + \Eta\{X_2\given X_1\} + \cdots+\Eta\{X_n \given X_1,\ldots,X_{n-1}\}.$$

Next we show that a constant random variable has zero entropy. This reflects the idea that we get
no information from sampling a random variable if it always takes the same value no matter what.

@@

\proof First suppose that $X$ takes only one value.
Let $a$ be the value of $X$ such that $\pr\{X=a\} = 1$.
Then $(X,X)$ equals $(a,a)$ with probability $1$ as well,
so $\Eta\{X\} = \Eta\{X,X\}$ by the invariance axiom. But it can easily be checked that
$X$ and $(X,X)$ are independent (we have, for example, $\pr\bigl\{X = a, (X,X)= (a,a)\bigr\} =
\pr\{X = a\}\pr\bigl\{(X,X) = (a,a)\bigr\}$), so $\Eta\{X,X\} = 2\Eta\{X\}$. Thus we conclude
that $\Eta\{X\} = 0$.

Now suppose that $X$ takes more than one value; let $A$ be the set of $a$ such that
$\pr\{X = a\} > 0$.
\slug

Now we establish the intuitive fact
that the entropy of a uniform random variable supported on a set $A$ is at most
the entropy of a uniform random variable supported on a superset $B$ of $A$.

\proclaim Proposition \advthm. Let $A\subseteq B$ with $B$ finite,
let $X$ be uniformly distributed on $A$, and let $Y$ be uniformly distributed on $B$.
Then $\Eta\{X\} \le \Eta\{Y\}$, with equality if and only if $A = B$.

\proof By the extensibility axiom, $\Eta\{X\}$ is not affected if we regard $X$ as a function
taking values in $B$. Then by the maximisation axiom, $\eta\{X\}\le \eta\{Y\}$, since $Y$ is
uniform on $B$. If $A = B$, then it is clear that $\Eta\{X\} = \Eta\{Y\}$, since $X$ and $Y$
are the same random variable.
Now say $|A| = m$ and $|B| = n$ with $m<n$. If $m=1$, then by the previous proposition
we have $\Eta\{X\} = 0$, and $Y$ takes at least two values,

The next proposition is that $\Eta\{X\}$ is always nonnegative. The following
proof is attributed to S.~Prendiville.

\proclaim Proposition \advthm. For any random variable $X$ taking values in a finite set $A$,
we have $\Eta\{X\} \ge 0$.

\proof First we consider the case that there exists $n\in \NN$ such that for all $x\in A$, there is some
$c_x$ such that $\pr\{X = x\} = c_x/n$. Let $Y$ be uniform on $[n]$ and let $E_x$ be a partition
of $[n]$ such that $|E_x| = c_x$ for all $x\in X$. Then let $Z$ be given by $Z = x$ if $Y\in E_x$.
This random variable $Z$ takes values in $A$ and has the exact same distribution as $X$, so
by the invariance axiom, $\Eta\{X\} = \Eta\{Z\}$.
This means that $Z = f(Y)$ for some function $f$, meaning that there is a bijection between
values $y$ taken by $Y$ and values $\bigl(y, f(y)\bigr)$ taken by $(Y,Z)$. So we may apply the invariance
axiom again to conclude that $\Eta\{Y,Z\} = \Eta\{Y\}$.

Now by the additivity axiom, $\Eta\{Y,Z\} = \Eta\{Y\given Z\} + \Eta\{Z\}$, so we
have $\Eta\{Y\} = \Eta\{Y\given Z\} + \Eta\{Z\}$.

If $X$ is a random variable such that $X = f(X)$ for some random variable $Y$, then we say
that $X$ is {\it determined by} $Y$ or $Y$ {\it determines} $X$. The following proposition formalises
the idea that we get more information from $Y$ than we get from $X$.

\proclaim Proposition \advthm. Let $X$ and $Y$ be random variables such that $Y$ determines $X$. Then
$$\Eta\{ X\} \le \Eta\{Y\}.$$

\proof The

For random variables $X$ and $Y$, the {\it mutual information} $\II\{X : Y\}$ is defined
by the equivalent formulas
$$\eqalign{
\II\{ X : Y\} &= \Eta\{X\} + \Eta\{Y\} - \Eta\{X,Y\} \cr
&= \Eta\{X\} - \Eta\{X\given Y\} \cr
&= \Eta\{Y\} - \Eta\{Y\given X\}. \cr
}$$
It measures, roughly speaking, how much information one can get from one variable
by looking at the other one. From the formula it is clear that $\II\{X : Y\}  = 0$
if and only if $X$ and $Y$ are independent, and we also have the
inequality $\Eta\{X \given Y\} \le \Eta\{X\}$. Given a triple $(X,Y,Z)$ of random variables,
we can apply this with $(X\given Z = z)$ to obtain the inequality
$$ \Eta\{ X\given Y,Z\} \le \Eta\{X\given Z\},$$
which is called {\it submodularity}. An equivalent statement is
$$\Eta\{X,Y,Z\} + \Eta\{Z\} \le \Eta\{X,Z\} + \Eta\{Y,Z\}.$$
If $Z$ takes values in a set $C$, the {\it conditional mutual information} is defined by
$$\eqalign{
\II\{ X : Y \given Z\} &= \sum_{z\in C} p_Z(z) \II\bigl\{ (X\given Z = z) : (Y\given Z = z)\bigr\} \cr
&= \Eta\{ X\given Z\} + \Eta\{Y\given Z\} - \Eta\{ X, Y \given Z \} \cr
&= \Eta\{ X, Z\} -2\Eta\{ Z\} + \Eta\{Y,Z\} - \Eta\{ X, Y ,Z \} + \Eta\{Z\}\cr
&= \Eta\{ X, Z\} + \Eta\{Y,Z\} - \Eta\{ X, Y ,Z \} - \Eta\{Z\},\cr
}$$
so we see that submodularity is equivalent to the statement $\II\{X : Y\given Z\} \ge 0$.
Analogously to the unconditional case, we have equality if and only if $X$ and $Y$ are
independent when conditioned on $Z$.

\medskip\boldlabel The formula for entropy.
Now we finally give the formula for the entropy of a random variable. For a discrete random
variable $X$, we let $p_X : S\to [0,1]$ denote the probability mass function of $X$, given by
$p_X(x) = \pr\{X = x\}$ for all $x\in S$. Unless subscripted, $\log$ always denotes the natural logarithm.

\proclaim Proposition \advthm. Let $X$ be a random variable taking values in a finite set $S$.
Then $\Eta\{X\}$ satisfies axioms (a) through (e) as well as the additional normalisation axiom (f)
if and only if
$$\Eta\{X\} = \ex\bigl\{ 1/\log p_X(X) \bigr\} = \sum_{x\in S} p_X(x) \log {1\over p_X(s)},$$
where we adopt the convention that $0\log(1/0) = 0$.

\proof [TODO: Prove this.]\slug

Equipped now with a formula for the entropy $\Eta\{X\}$ of a random variable $X$, we can prove further useful
facts. We saw in the previous section that the entropy of a uniform random variable whose range is $\{0,1\}^n$
is $\log_2(e)\cdot n = \log(2^n)$, and we know, by the maximisation axiom, that this is the largest
the entropy of any $X$ taking values in $\{0,1\}^n$ can be. The following proposition generalises
this notion.

\proclaim Proposition \advthm. For any random variable $X$ taking values in a finite set $S$,
we have
$$\Eta\{X\} \le \log |S|.$$

\proof By Jensen's inequality, we have
$$\eqalign{
\Eta\{X\} &= \ex\bigl\{ \log(1/p_X(X)\bigr\} \cr
&\le \log \ex\bigl\{ 1/p_X(X)\bigr\} \cr
&= \log \sum_{x\in S} \pr\{X = x\} {1\over \pr\{X = x\}} \cr
&= \log |S|.\noskipslug \cr
}$$

>>>>

\def\eps{\epsilon}
\def\FF{{\bf F}}
\def\bar#1{\overline{#1}}
\def\hat#1{\widehat{#1}}
\def\norm#1{|\!|#1|\!|}
\def\bignorm#1{\big|\!\big|#1\big|\!\big|}
\def\Norm#1{\Big|\!\Big|#1\Big|\!\Big|}
\def\normm#1{\bigg|\!\bigg|#1\bigg|\!\bigg|}
\def\Bohr{\op{\rm Bohr}}
\def\Eta{\op{\bf H}}
\def\II{\op{\bf I}}
\def\dd{\op{\bf d}}
\def\given{\mathbin{|}}

\widemargins
\bookheader{ENTROPY AND ADDITIVE COMBINATORICS}{MARCEL K. GOH}

\maketitle{Entropy and additive combinatorics}{by}{Marcel K. Goh}{16 January 2024}

@@

\proclaim Proposition \advthm. Let $X$ and $Y$ be independent random variables. Then
$\Eta\{Y\given X\} = \Eta\{Y\}$ and consequently
$\Eta\{X,Y\} = \Eta\{X\} + \Eta\{Y\}$.

\proof Suppose $X$ takes values in a finite set $A$. Then for all $x\in A$, the distribution of
$Y$ and $Y$ given that $X = x$ is the same, so
$$\Eta\{Y \given X\} = \sum_{x\in A} \pr\{X = x\} \Eta\{Y\given X = x\} = \sum_{x\in A} \pr\{X = x\} \Eta\{Y\}
=\Eta\{Y\}.$$
The second version of the statement follows from the additivity axiom.\slug

@@

\edef\corxn{\the\thmcount}
\proclaim Corollary \advthm. For a discrete random variable $X$, we have $\Eta\{X^n\} = n\Eta\{X\}$.

@@

In the case where $X$ is the uniform distribution on $\{0,1\}$, and $X_1,\ldots,X_n$ are independent
and distributed as $X$,
we can additionally use the normalisation axiom to conclude that
$$\Eta\{X^n\} = \Eta\{X_1,\ldots,X_n\} = \sum_{i=1}^n \Eta\{X_i\} = \log_2(e)\cdot n,$$
by induction.

A similar induction on general random variables $X_1,\ldots,X_n$ gives the following statement,
often known as the {\it chain rule}.

\parenproclaim Proposition~{\advthm} (Chain rule).
Let $X_1,\ldots,X_n$ be random variables. Then
$$\Eta\{X_1,\ldots,X_n\} = \Eta\{X_1\} + \Eta\{X_2\given X_1\}
+ \cdots+\Eta\{X_n \given X_1,\ldots,X_{n-1}\}.\noskipslug$$

Next we establish the intuitive fact
that the entropy of a uniform random variable supported on a set $A$ is at most
the entropy of a uniform random variable supported on a superset $B$ of $A$.

\edef\propunifineq{\the\thmcount}
\proclaim Proposition \advthm. Let $A\subseteq B$ with $B$ finite,
let $X$ be uniformly distributed on $A$, and let $Y$ be uniformly distributed on $B$.
Then $\Eta\{X\} \le \Eta\{Y\}$, with equality if and only if $A = B$.

\proof By the extensibility axiom, $\Eta\{X\}$ is not affected if we regard $X$ as a function
taking values in $B$. Then by the maximisation axiom, $\Eta\{X\}\le \Eta\{Y\}$, since $Y$ is
uniform on $B$. If $A = B$, then it is clear that $\Eta\{X\} = \Eta\{Y\}$, since $X$ and $Y$
are the same random variable.
On the other hand, say $|A| = m$ and $|B| = n$ with $m<n$. If $m=1$, then by the previous proposition
we have $\Eta\{X\} = 0$, and by normalisation and invariance, $\Eta\{Y\} = \log_2(e)$.\slug

If $Y$ is a random variable such that $Y = f(X)$ for some random variable $Y$ and some function $f$,
then we say that $Y$ is {\it determined by} $X$ or $X$ {\it determines} $Y$.
We want to show that $\Eta\{Y\}\le \Eta\{X\}$, which reflects the idea that we get more information
from $X$ than from $Y$. This, rather annoyingly, seems to require two steps.

\proclaim Lemma \advthm. If $Y = f(X)$ then $\Eta\{X\} = \Eta\{Y\} + \Eta\{X\given Y\}$.

\proof There is a a bijection between values $x$ taken
by $Y$ and values $\bigl(x,f(x)\bigr)$ taken by $(X,Y)$, so we have
$$\Eta\{X\} = \Eta\{X,Y\} = \Eta\{Y\} + \Eta\{X\given Y\}$$
by additivity.\slug

We are now done if we can show that entropy is nonnegative.
This can also be derived straight from the axioms. The following proof is due to S.~Eberhard.

\proclaim Proposition \advthm. Let $X$ be a discrete random variable taking values in a finite
set $A$. Then $\Eta\{X\} \ge 0$.

\proof First we will work in the case where there exists $n$ such that $\pr\{X = a\}$ is a multiple of
$1/n$ for all $a\in A$. Let $Y$ be uniformly distributed on $[n]$ and let $\{E_a\}_{a\in A}$
be a partition of $[n]$ such that $|E_a| = n\pr\{X= a\}$ for all $a\in A$, and let $Z = a$
if $Y\in E_a$. This definition makes $Z$ and $X$ identically distributed, so $\Eta\{Z\} = \Eta\{X\}$
by the invariance axiom, and it suffices to prove $\Eta\{Z\} \ge 0$.

Since $Z$ is determined by $Y$, we have $\Eta\{Y\} = \Eta\{Z\} + \Eta\{Y\given Z\}$ by the previous lemma.
Furthermore, for every $a\in A$, the conditional entropy $\Eta\{Y \given Z = a\}$ is uniformly
distributed on a set of size at most $n$, whereas is uniformly distributed on $n$, so by
Proposition~{\propunifineq}, we have $\pr\{Y \given Z = a\} \le \Eta\{Y\}$. This implies
that $\Eta\{Y\} \ge \Eta\{Y\given Z\}$, which in turn gives us $\Eta\{Z\}\ge 0$.\slug

\proclaim Corollary \advthm. If $Y = f(X)$ then $\Eta\{X\} \ge \Eta\{Y\}$.\slug

Now we show that a random variable has zero entropy if and only if it is constant.
This reflects the idea that the variables from which we get no information are those which take
the same value no matter what.

@@

\proof First suppose that $X$ takes only one value.
Let $a$ be the value of $X$ such that $\pr\{X=a\} = 1$.
Then $(X,X)$ equals $(a,a)$ with probability $1$ as well,
so $\Eta\{X\} = \Eta\{X,X\}$ by the invariance axiom. But it can easily be checked that
$X$ and $(X,X)$ are independent (we have
$$\pr\bigl\{X = a, (X,X)= (a,a)\bigr\} = \pr\{X = a\}\pr\bigl\{(X,X) = (a,a)\bigr\}$$
for instance), so $\Eta\{X,X\} = 2\Eta\{X\}$. Thus we conclude
that $\Eta\{X\} = 0$.

Now suppose that $X$ takes more than one value; let $A$ be the set of $a$ such that
$\pr\{X = a\} > 0$ and let $\alpha = \max_{a\in A} \pr\{X = a\}$. For all $n$ let $X^n$ denote
the tuple of $n$ independent copies of $X$; the maximum probability of any particular value (in $A^n$)
that $X^n$ takes is $\alpha^n$. But $\alpha < 1$ since $X$ takes more than one value, so for any
$\eps > 0$ we can find $n$ such that $\alpha^n < \eps$. This means that we can partition $A^n$ into
two disjoint sets $E$ and $F$ such that $\pr\{X^n \in E\}$ and $\pr\{X^n \in F\}$ are both
in the range $[1/2-\eps, 1/2+\eps]$.

Let $Y$ be the random variable taking the value $0$ if $X^n\in E$ and $1$ if $X^n\in F$. Then
by Corollary~{\corxn}, $\Eta\{X^n\} = n\Eta\{X\}$, and since $X^n$ determines $Y$,
$$\Eta\{X^n\} = \Eta\{Y\} + \Eta\{X^n\given Y\} \ge \Eta\{Y\}.$$
But $\Eta\{Y\} > 0$ for $\eps$ small enough, the normalisation and continuity axioms. So
$\Eta\{X\} \ge \Eta\{Y\}/n > 0$ as well.\slug

\medskip\boldlabel Mutual information.
For random variables $X$ and $Y$, the {\it mutual information} $\II\{X : Y\}$ is defined
by the equivalent formulas
$$\eqalign{
\II\{ X : Y\} &= \Eta\{X\} + \Eta\{Y\} - \Eta\{X,Y\} \cr
&= \Eta\{X\} - \Eta\{X\given Y\} \cr
&= \Eta\{Y\} - \Eta\{Y\given X\}. \cr
}$$
It measures, roughly speaking, how much information one can get from one variable
by looking at the other one. From the formula it is clear that $\II\{X : Y\}  = 0$
if and only if $X$ and $Y$ are independent, and we also have the
inequality $\Eta\{X \given Y\} \le \Eta\{X\}$; that is, {\sl conditioning cannot increase the
entropy of a random variable}. Given a triple $(X,Y,Z)$ of random variables,
we can apply this with $(X\given Z = z)$ to obtain the inequality
$$ \Eta\{ X\given Y,Z\} \le \Eta\{X\given Z\},$$
which is called {\it submodularity}. An equivalent statement is
$$\Eta\{X,Y,Z\} + \Eta\{Z\} \le \Eta\{X,Z\} + \Eta\{Y,Z\}.$$
If $Z$ takes values in a set $C$, the {\it conditional mutual information} is defined by
$$\eqalign{
\II\{ X : Y \given Z\} &= \sum_{z\in C} p_Z(z) \II\bigl\{ (X\given Z = z) : (Y\given Z = z)\bigr\} \cr
&= \Eta\{ X\given Z\} + \Eta\{Y\given Z\} - \Eta\{ X, Y \given Z \} \cr
&= \Eta\{ X, Z\} -2\Eta\{ Z\} + \Eta\{Y,Z\} - \Eta\{ X, Y ,Z \} + \Eta\{Z\}\cr
&= \Eta\{ X, Z\} + \Eta\{Y,Z\} - \Eta\{ X, Y ,Z \} - \Eta\{Z\},\cr
}$$
so we see that submodularity of entropy is equivalent to the statement that $\II\{X : Y\given Z\} \ge 0$.
Analogously to the unconditional case, we have equality if and only if $X$ and $Y$ are
independent when conditioned on $Z$.

\advsect Group-valued random variables

Now we will examine the case where the random variables in question take values in an abelian group $G$,
meaning we can take sums $X+Y$ and differences $X-Y$ of them. Note that if we condition on $Y$,
then the values taken $X+Y$ are in bijection with values taken by $X$. This leads to the following
proposition.

\edef\maxsumsetbound{\the\thmcount}
\proclaim Proposition \advthm. Let $X$ and $Y$ be random variables each taking finitely
many values in an abelian group $G$.
We have
$$\max\bigl( \Eta\{X\}, \Eta\{Y\} \bigr) - \II\{X : Y\} \le \Eta\{ X \pm Y\}.$$
Furthermore, for any random variable $Z$, we have the conditional version
$$\max\bigl( \Eta\{X\given Z\}, \Eta\{Y\given Z\} \bigr) - \II\{X : Y\given Z\} \le \Eta\{ X \pm Y\given Z\}$$
of the same statement.

\proof Since conditioning does not increase entropy, we have
$$\Eta\{ X\pm Y\} \ge \Eta\{ X\pm Y \given Y\},$$
and since the probabilities $\pr\{ X + Y = z \given Y = y\} = \pr\{ X = z-y \given Y = y\}$ for all $z\in G$,
by invariance we have
$$\Eta\{ X\pm Y\} \ge \Eta\{X\given Y\} = \Eta\{X\} - \II\{X : Y\}.$$
Repeating the same argument but exchanging the roles of $X$ and $Y$, we get
$$\Eta\{ X\pm Y\} \ge \Eta\{Y\given X\} = \Eta\{Y\} - \II\{X : Y\},$$
so
$$\Eta\{ X\pm Y\} \ge \max\bigl(\Eta\{X\}, \Eta\{Y\}\bigr) - \II\{X : Y\}.$$
Now let $Z$ be any random variable with finite support.
$$\eqalign{
\Eta\{ X\pm Y\given Z\} &= \sum_{z\in G} \pr\{Z = z\} \Eta\{ X \pm Y \given Z = z\} \cr
&\ge \Bigl(\max\bigl(\Eta\{X\given Z\}, \Eta\{Y\given Z\}\bigr) - \II\{X : Y\given Z\} \Bigr)
\sum_{z\in G} \pr\{Z = z\} \cr
&= \max\bigl(\Eta\{X\given Z\}, \Eta\{Y\given Z\}\bigr) - \II\{X : Y\given Z\} ,\cr
}$$
which completes the proof.\slug

\proclaim Corollary \advthm. If $X$ and $Y$ are independent, then
$$\max\bigl( \Eta\{X\}, \Eta\{Y\} \bigr) \le \Eta\{ X \pm Y\}.$$

\proof The mutual information $\II\{X : Y\}$ is zero whenever $X$ and $Y$ are independent.\slug

\medskip\boldlabel Entropic Ruzsa distance.
In additive combinatorics, whenever we have two finite subsets $A$ and $B$
of the same abelian group, we can compute the Ruzsa distance
$$d(A,B) = \log {|A-B|\over \sqrt{|A|\cdot|B|}}$$
between them. (This satisfies all the axioms of a metric except the one requiring $d(A,A) = 0$ for all sets
$A$.)

The entropic analogue of the Ruzsa distance is defined as follows.
For finitely supported random variables $X$ and $Y$ taking values in the same abelian group,
we let $X'$ and $Y'$ be independent copies of $X$
and $Y$, respectively, and define the {\it entropic Ruzsa distance} by
$$\dd\{X,Y\} = \Eta\{X'-Y'\} - {\Eta\{X'\} \over 2} - {\Eta\{Y'\}\over 2}.$$
This definition only depends on the individual distributions of $X$ and $Y$ and does not require them to
have the same sample space. Once again, we don't necessarily have $\dd\{X,X\} = 0$, but we do have
the triangle inequality, which shall now prove.

\proclaim Proposition \advthm. Let $X$, $Y$, and $Z$ be random variables with finite support in the same
abelian group. Then
$$\dd\{X,Z\} \le \dd\{X,Y\} + \dd\{Y,Z\},$$
which is equivalent to
$$\Eta\{X'-Z'\} \le \Eta\{X'-Y'\} + \Eta\{Y'-Z'\} - \Eta\{Y'\}$$
for $X'$, $Y'$, and $Z'$ independent and distributed as $X$, $Y$, and $Z$, respectively.

\proof That the two statements are equivalent is easily obtained by expanding the definition of
entropic Ruzsa distance and cancelling some terms. So without loss of generality, we may assume that
that $X$, $Y$, and $Z$ are independent and just prove the second statement.

By submodularity, we have $\II\bigl\{ (X-Y : Z) \given X-Z\bigr\} \ge 0$, so
\edef\eqfirstsubmodularity{\the\eqcount}
$$\eqalign{
0 &\le \II\bigl\{ (X-Y : Z) \given X-Z\bigr\} \cr
&\le \Eta\{X-Y \given X-Z\} + \Eta\{ Z\given X-Z\} - \Eta\{ X-Y, Z \given X-Z\} \cr
&\le \Eta\{X-Y, X-Z\} + \Eta\{ Z, X-Z\} - \Eta\{ X-Y, Z, X-Z\} - \Eta\{X-Z\}.\cr
}\adveq$$
Now, since the values $(x-y,x-z)$ taken by $(X-Y,X-Z)$ are in bijection with values
$(x-z,y-z)$ taken by $(X-Z, Y-Z)$ via the map $(v,w) \mapsto (w, w-v)$,
by the invariance axiom we have
$$\Eta\{X-Y, X-Z\} = \Eta\{X-Z, Y-Z\},$$
and
$$\Eta\{X-Y, X-Z\} \le \Eta\{X-Y\} + \Eta\{Y-Z\}$$
follows by submodularity. Similar invocations of the invariance axiom give
$$\Eta\{Z, X-Z\} = \Eta\{X,Z\}$$
and
$$\Eta\{X-Y, Z, X-Z\} = \Eta\{X,Y,Z\} = \Eta\{X,Z\} + \Eta\{Y\},$$
where in the latter statement the second equality follows from the fact that $(X,Y)$ and $Z$ are
independent.
Substituting these three inequalities into~\refeq{\eqfirstsubmodularity}, we have
$$0\le \Eta\{X-Y\} + \Eta\{Y-Z\} +\Eta\{X,Z\} -\Eta\{X,Z\} + \Eta\{Y\} - \Eta\{X-Z\},$$
whence
$$ \Eta\{X-Z\} \le \Eta\{X-Y\} + \Eta\{Y-Z\} + \Eta\{Y\},$$
which completes the proof.\slug

We also define a conditional version of the entropic Ruzsa distance. If $X$ and $Y$ are $G$-valued
random variables with finite support and $Z$ and $W$ are any random variables with finite supports
$A$ and $B$ respectively, then we define
$$\dd\{ X\given Z ; Y\given W\} = \sum_{z\in A} \sum_{w\in B} \pr\{Z = z\} \pr\{W=w\}
\dd\bigl\{ (X\given Z = z) ; (Y\given W=w)\bigr\}.$$
If $(X',Z')$ and $(Y',W')$ are independent copies of $(X,Z)$ and $(Y,W)$ respectively, then this
distance is also given by the formula
$$\dd\{ X\given Z ; Y\given W\} = \Eta\{X'-Y'\given Z',W'\} - {\Eta\{X'\given Z'\}\over 2}
- {\Eta\{Y'\given W'\}\over 2}.$$

\advsect The Pl\"unnecke--Ruzsa inequality

In additive combinatorics, one of the most useful sumset inequalities is the following.

\parenproclaim Theorem A (Pl\"unnecke--Ruzsa inequality). Let $A$ and $B$ be finite subsets of an
abelian group and suppose that $|A+B|\le K|A|$ for some constant $K$. Then for any integers
$r,s\ge 0$, not both zero, we have $|rB-sB| \le K^{r+s}|A|$.\slug

In this section we will develop an entropic analogue of this statement, in which sets are replaced
by random variables of finite support and cardinality is replaced with the exponential of
entropy. First, a technical lemma.

\proclaim Lemma \advthm. Let $X$, $Y$, and $Z$ be independent random variables taking values in a common
abelian group. Then
$$\Eta\{X+Y+Z\} - \Eta\{X+Y\} \le \Eta\{Y+Z\} -\Eta\{Y\}.$$

\proof By submodularity, the quantity $\II\{ X : Z\given X+Y+Z\}$ is nonnegative, so we have
$$\eqalign{
0&\le \II\{ X : Z\given X+Y+Z\} \cr
&= \Eta\{X, X+Y+Z\} + \Eta\{Z, X+Y+Z\}\cr
&\qquad\qquad\qquad\qquad- \Eta\{X,Z,X+Y+Z\} - \Eta\{X+Y+Z\}.\cr
}$$
Since $X$, $Y$, and $Z$ are independent, we have
$$\Eta\{X,X+Y+Z\} = \Eta\{X,Y+Z\} = \Eta\{X\} + \Eta\{Y+Z\},$$
where in the first equality we use invariance. By similar reasoning we have
$$\Eta\{Z, X+Y+Z\} = \Eta\{Z\} + \Eta\{X+Y\}$$
and
$$\Eta\{X,Z,X+Y+Z\} = \Eta\{X\} + \Eta\{Y\} + \Eta\{Z\}.$$
Plugging these three identities into the inequality above yields
$$\eqalign{
0&\le \Eta\{X\} + \Eta\{Y+Z\} + \Eta\{Z\} + \Eta\{X+Y\} \cr
&\qquad\qquad -\Eta\{X\} - \Eta\{Y\} - \Eta\{Z\} - \Eta\{X+Y+Z\} \cr
&= \Eta\{Y+Z\} + \Eta\{X+Y\} - \Eta\{Z\} - \Eta\{X+Y+Z\},\cr
}$$
whence the claim follows upon rearranging.\slug

From here we are not far from proving the entropic Pl\"unnecke--Ruzsa inequality, a result of T.~Tao.

\proclaim Theorem \advthm. Let $X,Y_1,\ldots,Y_m$ be independent random variables of finite entropy
taking values in an abelian group $G$, such that
$$\Eta\{X+Y_i\} \le \Eta\{X\} + \log K_i$$
for all $1\le i\le m$ and some scalars $K_1,\ldots,K_m\ge 1$. Then
$$\Eta\{X+Y_1+\cdots+Y_m\} \le \Eta\{X\} + \log(K_1\cdots K_m).$$

\proof We prove the claim by induction on $m$. If $m=1$, then we are done by hypothesis. Now
suppose that $\Eta\{X+Y_1+\cdots+Y_{m-1}\} \le \Eta\{X\} + \log(K_1\cdots K_{m-1})$. Then by
the previous lemma, the induction hypothesis, and the hypothesis on $\Eta\{X+Y_m\}$, we have
$$\eqalign{
\Eta\{Y_1 + \cdots + Y_{m-1} + X + Y_m\}
&\le \Eta\{Y_1 + \cdots + Y_{m-1} + X\} \cr
&\qquad\qquad\qquad\qquad+ \Eta\{X+Y_m\} - \Eta\{X\}\cr
&\le \Eta\{X\} + \log(K_1\cdots K_{m-1}) + \log K_m\cr
&\le \Eta\{X\} + \log(K_1\cdots K_m),\cr
}$$
which is what we sought to prove.\slug

We can make this look bit more like the version of the Pl\"unnecke--Ruzsa inequality above
by using the triangle inequality.

\parenproclaim Corollary {\advthm} (Entropic Plunnecke--Ruzsa inequality). Let $X$ and $Y$ be random variables
with $\Eta\{X+Y\} \le \Eta\{X\} + \log K$. Then for any $r,s\ge 0$ not both zero, we have
$$\Eta\{ Y_1 + \cdots + Y_r - Z_1 - \cdots - Z_s \} \le \Eta\{X\} + (r+s)\log K,$$
where $Y_1,\ldots,Y_r,Z_1,\ldots,Z_s$ are independent copies of $Y$.

\proof By the entropic Ruzsa triangle inequality, we have
$$\eqalign{
&\Eta\{ Y_1 + \cdots + Y_r - Z_1 - \cdots - Z_s \} \le \cr
&\qquad\qquad\Eta\{Y_1 + \cdots + Y_r + X\} + \Eta\{-X - Z_1 - \cdots - Z_s\} - \Eta\{-X\}. \cr
}$$
The values of $-X$ are in bijection with values of $X$, and the values of $-X-Z_1-\cdots-Z_s$
are in bijection with the values of $X+Z_1+\cdots+Z_s$ (with the same probabilities in both cases),
so by the invariance axiom, we have
$$\eqalign{
&\Eta\{ Y_1 + \cdots + Y_r - Z_1 - \cdots - Z_s \} \le \cr
&\qquad\qquad\Eta\{X+Y_1 + \cdots + Y_r\} + \Eta\{X+ Z_1 + \cdots + Z_s\} - \Eta\{X\},\cr
}$$
and we can apply the the previous theorem twice to get
$$\eqalign{
\Eta\{ Y_1 + \cdots + Y_r - Z_1 - \cdots - Z_s \}  &\le \Eta\{X\} + \log (K^r) + \log (K^s),\cr
&= \Eta\{X\} + (r+s)\log K.\noskipslug\cr
}$$

% \medskip\boldlabel The formula for entropy.
% Now we finally give the formula for the entropy of a random variable. For a discrete random
% variable $X$, we let $p_X : S\to [0,1]$ denote the probability mass function of $X$, given by
% $p_X(x) = \pr\{X = x\}$ for all $x\in S$. Unless subscripted, $\log$ always denotes the natural logarithm.
% 
% \proclaim Proposition \advthm. Let $X$ be a random variable taking values in a finite set $S$.
% Then $\Eta\{X\}$ satisfies axioms (a) through (e) as well as the additional normalisation axiom (f)
% if and only if
% $$\Eta\{X\} = \ex\bigl\{ 1/\log p_X(X) \bigr\} = \sum_{x\in S} p_X(x) \log {1\over p_X(s)},$$
% where we adopt the convention that $0\log(1/0) = 0$.
% 
% \proof [TODO: Prove this.]\slug
% 
% Equipped now with a formula for the entropy $\Eta\{X\}$ of a random variable $X$, we can prove further useful
% facts. We saw in the previous section that the entropy of a uniform random variable whose range is $\{0,1\}^n$
% is $\log_2(e)\cdot n = \log(2^n)$, and we know, by the maximisation axiom, that this is the largest
% the entropy of any $X$ taking values in $\{0,1\}^n$ can be. The following proposition generalises
% this notion.
% 
% \proclaim Proposition \advthm. For any random variable $X$ taking values in a finite set $S$,
% we have
% $$\Eta\{X\} \le \log |S|.$$
% 
% \proof By Jensen's inequality, we have
% $$\eqalign{
% \Eta\{X\} &= \ex\bigl\{ \log(1/p_X(X)\bigr\} \cr
% &\le \log \ex\bigl\{ 1/p_X(X)\bigr\} \cr
% &= \log \sum_{x\in S} \pr\{X = x\} {1\over \pr\{X = x\}} \cr
% &= \log |S|.\noskipslug \cr
% }$$
